---
title: "Adding tables and figures"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>",
                      message = FALSE, warning = FALSE)
```

```{r setup}
library(rtfreporter)
```

A rtfreporter document is a **flat sequence of pages**, and the rule is simple:
**one page holds exactly one content item** -- a table or a figure -- plus an
optional title and footnote block. You add content with `rtf_tables()` (and
`rtf_figures()` for images); each element you pass becomes one page.

This article is the deep dive on content. For document-wide settings (page size,
fonts) see [Page & document setup](page-setup.html); for converting `gt` /
`gtsummary` / `rtables` objects see
[Importing tables](importing-tables.html); for splitting a long table across
pages see [Paginating with as_rtftables()](pagination.html).

```{r data}
df1 <- data.frame(Parameter = c("Age", "  Mean", "  SD"),
                  Value     = c("", "75.1", "8.2"), stringsAsFactors = FALSE)
df2 <- data.frame(Parameter = c("Sex", "  F", "  M"),
                  Value     = c("", "53%", "47%"), stringsAsFactors = FALSE)
```


## A single table

The quickest way is to hand a data.frame straight to `rtf_tables()` with the
formatting you want -- a single item needs no `list()`:

```{r single-bare}
doc <- rtf_document() |>
  rtf_tables(df1, border = "tfl", row_height_twips = 280L, col_rel_width = c(2, 1))
length(doc$contents)        # one page
```

For full control, build the table object yourself with `rtftable()` (or
`as_rtftables()` for gt/rtables sources) and pass that:

```{r single-built}
tbl <- rtftable(df1, border = "tfl", col_rel_width = c(2, 1),
                col_spec = list(list(col = 2, align = "right")))
doc <- rtf_document() |> rtf_tables(tbl)
```

The formatting arguments of `rtf_tables()` (`border`, `col_rel_width`,
`row_height_twips`, `col_spec`, ...) are documented on `?rtftable`; they are the
same options either way.


## The same settings for several tables

Pass a **list** of bare data.frames and the formatting arguments are applied to
**every** one -- handy for a batch of tables that should look identical. Each
element still becomes its own page:

```{r bulk}
doc <- rtf_document() |>
  rtf_tables(
    list(df1, df2),
    border           = "tfl",
    row_height_twips = 280L,
    col_rel_width    = c(2, 1)
  )
length(doc$contents)        # two pages, same look
```


## Different settings per table

When each table needs its own look, build each one separately (with its own
settings) and pass the list. Pre-built `rtftable()` objects keep their own
settings:

```{r individual}
t1 <- rtftable(df1, border = "tfl", row_height_twips = 280L)
t2 <- rtftable(df2, border = NULL,  row_height_twips = 200L)   # no borders

doc <- rtf_document() |> rtf_tables(list(t1, t2))
doc$contents[[1]]$row_height_twips   # 280 -- t1 keeps its own
doc$contents[[2]]$row_height_twips   # 200 -- t2 keeps its own
```

### Override semantics

A formatting argument you pass **explicitly** to `rtf_tables()` overrides that
field on every pre-built table; arguments you leave at their default do not
touch the tables'' own values. This lets you set a shared override without
rebuilding the tables:

```{r override}
doc <- rtf_document() |>
  rtf_tables(list(t1, t2), row_height_twips = 240L)   # explicit -> overrides both
doc$contents[[1]]$row_height_twips   # 240
doc$contents[[2]]$row_height_twips   # 240
```

(Bare data.frames are always built from the supplied arguments; `rtfplot()`
figure objects are never modified.)


## Titles and footnotes

Each page can carry a title block and a footnote block. Pass `titles` /
`footnotes` as lists parallel to the content (one element per page; each a
character vector, one line per element). Use `NULL` for a page with none:

```{r titles}
doc <- rtf_document() |>
  rtf_tables(
    list(df1, df2),
    border    = "tfl",
    titles    = list(c("Table 14.1.1", "Demographics"),
                     c("Table 14.1.2", "Sex")),
    footnotes = list(c("Source: ADSL"), NULL)
  )
```

You can also set them after the content with `rtf_titles()` / `rtf_footnotes()`.
(When the content comes from a `gt` table via `as_rtftables()`, its title and
source notes flow through automatically -- see
[Importing tables](importing-tables.html).)


## Figures

Images are content too: one figure per page, added with `rtf_figures()`. Pass a
list of image paths (PNG/JPEG) or `rtfplot()` objects; `width_twips` /
`height_twips` / `align` control the display, and `titles` / `footnotes` work
exactly as for tables.

```{r figure}
# A small example PNG to embed.
img <- tempfile(fileext = ".png")
png(img, width = 480, height = 320); plot(1:10, main = "Mean over time"); dev.off()

doc <- rtf_document() |>
  rtf_figures(list(img), width_twips = 6000L, align = "center",
              titles = list(c("Figure 14.2.1", "Mean over time")))
length(doc$contents)        # one figure page
```

Tables and figures share the same page model, so a report is just a sequence of
content pages -- some tables, some figures -- each optionally titled and
footnoted. To control the running header/footer across those pages, see the
headers & footers article.

See `?rtf_tables` and `?rtf_figures` for the full argument reference.
