---
title: "TLG cookbook: pharmaverse tables -> RTF"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
# This article runs only when every table-building package is available.
.pkgs <- c("gtsummary", "rtables", "tern", "tfrmt", "dplyr")
.ok <- all(vapply(.pkgs, requireNamespace, logical(1L), quietly = TRUE))
knitr::opts_chunk$set(eval = .ok)
```

```{r, eval = TRUE, echo = FALSE, results = "asis"}
if (!.ok)
  cat("> This article renders only when **gtsummary, rtables, tern, tfrmt**",
      "and **dplyr** are installed.  The code below shows the workflow either",
      "way.\n")
```

rtfreporter does not compute statistics or format numbers -- the
[pharmaverse](https://pharmaverse.org) ecosystem already does that
beautifully.  rtfreporter handles the **last mile**: taking whatever table
object your team produced and laying it out as a clinical-style RTF page.

This cookbook mirrors the layout-by-example style of the
[pharmaverse TLG examples](https://pharmaverse.github.io/examples/tlg/) and
the [NEST TLG catalog](https://insightsengineering.github.io/tlg-catalog/stable/):
each entry **builds a table with a pharmaverse tool**, then **converts it to
RTF** with [`as_rtftables()`](../reference/as_rtftables.html) +
[`rtf_tables()`](../reference/rtf_tables.html) +
[`generate_rtfreport()`](../reference/generate_rtfreport.html).

```{r setup, eval = TRUE}
library(rtfreporter)
```

```{r libs}
library(gtsummary)
library(rtables)
library(tern)
library(tfrmt)
library(dplyr)
```

## Self-contained example data

To keep the article reproducible with no data dependency, we simulate a
small CDISC-style subject-level (`adsl`) and adverse-event (`adae`)
data set.  Swap in `pharmaverseadam` / `random.cdisc.data` in real work.

```{r data}
set.seed(2026)
n   <- 60
arm <- sample(c("Placebo", "Xanomeline Low", "Xanomeline High"), n, TRUE)

adsl <- tibble(
  USUBJID = sprintf("01-701-%03d", seq_len(n)),
  ARM     = factor(arm,
                   levels = c("Placebo", "Xanomeline Low", "Xanomeline High")),
  AGE     = round(rnorm(n, 60, 9)),
  SEX     = factor(sample(c("F", "M"), n, TRUE)),
  RACE    = factor(sample(c("WHITE", "BLACK OR AFRICAN AMERICAN", "ASIAN"),
                          n, TRUE, prob = c(.7, .2, .1)))
)

soc <- c("Gastrointestinal disorders", "Nervous system disorders",
         "Skin and subcutaneous tissue disorders")
pt  <- list("Nausea", "Diarrhoea", c("Headache", "Dizziness"),
            c("Rash", "Pruritus"))
adae <- adsl |>
  slice_sample(n = 90, replace = TRUE) |>
  transmute(
    USUBJID, ARM,
    AESOC   = factor(sample(soc, n(), TRUE)),
    AEDECOD = factor(sample(unlist(pt), n(), TRUE))
  )
```

A reusable clinical header / footer for every page:

```{r hf}
make_header <- function(table_no, title2) {
  rtf_header(rows = list(
    c(l = "Hoge Co. Limited",   r = "CONFIDENTIAL"),
    c(l = "Protocol: RTF-101",  r = "Page {AUTO_PAGE} of {AUTO_TOTAL_PAGES}"),
    c(c = paste0("Table ", table_no, "  ", title2))
  ))
}

# A single-column, left-aligned footer that records the table's provenance:
# the reference catalog, the source object type, and the package used.
make_footer <- function(source_object, built_with) {
  rtf_footer(c(l = paste0(
    "Layout adapted from the pharmaverse / NEST TLG catalogs.  ",
    "Source object: ", source_object, " (built with ", built_with, "); ",
    "read into RTF with rtfreporter::as_rtftables().  Data simulated.")))
}
```

## 1. Demographics & baseline (gtsummary)

A standard demographics table built with `gtsummary::tbl_summary()`:

```{r dm-build}
dm_tbl <- adsl |>
  select(ARM, AGE, SEX, RACE) |>
  tbl_summary(by = ARM,
              label = list(AGE = "Age (years)", SEX = "Sex",
                           RACE = "Race")) |>
  modify_header(label = "Characteristic")
dm_tbl
```

Convert it and write the RTF -- column labels, the by-group column
headers and the row structure all carry through:

```{r dm-rtf}
doc <- rtf_document() |>
  rtf_section(page = 1,
              secinfo = list(header = make_header("14.1.1",
                                                  "Demographic and Baseline Characteristics"),
                             footer = make_footer("gt_tbl", "gtsummary::tbl_summary()"))) |>
  rtf_tables(as_rtftables(dm_tbl, align_count_pct = TRUE),
             titles = list(c("Demographic and Baseline Characteristics",
                             "Safety Analysis Set")))

out <- tempfile(fileext = ".rtf")
generate_rtfreport(doc, out, overwrite = TRUE)
file.exists(out)
```

## 2. Adverse events by SOC / PT (tern + rtables)

The NEST stack (`tern` layouts on top of `rtables`) is the canonical way
to build a nested AE table:

```{r ae-build}
ae_lyt <- basic_table(show_colcounts = TRUE) |>
  split_cols_by("ARM") |>
  analyze_num_patients(vars = "USUBJID",
                       .stats = "unique",
                       .labels = c(unique = "Subjects with >=1 AE")) |>
  split_rows_by("AESOC", label_pos = "topleft", split_label = "SOC") |>
  count_occurrences(vars = "AEDECOD")

ae_tbl <- build_table(ae_lyt, as.data.frame(adae), alt_counts_df = as.data.frame(adsl))
ae_tbl
```

`as_rtftables()` reads the spanning column headers, the SOC row groups
(rendered with indentation) and the alignment straight from the
`VTableTree`:

```{r ae-rtf}
doc <- rtf_document() |>
  rtf_section(page = 1,
              secinfo = list(header = make_header("14.3.1",
                                                  "Adverse Events by SOC and Preferred Term"),
                             footer = make_footer("rtables TableTree", "tern + rtables"))) |>
  rtf_tables(as_rtftables(ae_tbl, align_count_pct = TRUE))

out <- tempfile(fileext = ".rtf")
generate_rtfreport(doc, out, overwrite = TRUE)
file.exists(out)
```

## 3. A pre-formatted table (tfrmt)

`tfrmt` describes the *format* of a table declaratively, then renders to
gt.  rtfreporter reads the gt it produces:

```{r tfrmt-build}
age_df <- adsl |>
  group_by(ARM) |>
  summarise(mean = mean(AGE), sd = sd(AGE), .groups = "drop") |>
  tidyr::pivot_longer(c(mean, sd), names_to = "param", values_to = "value") |>
  mutate(group = "Age (years)",
         label = recode(param, mean = "Mean", sd = "SD"),
         column = as.character(ARM))

tf <- tfrmt(
  group = group, label = label, column = column, param = param, value = value,
  body_plan = body_plan(
    frmt_structure(group_val = ".default", label_val = ".default", frmt("xx.x")))
)
tf_gt <- print_to_gt(tf, age_df)
tf_gt
```

```{r tfrmt-rtf}
doc <- rtf_document() |>
  rtf_section(page = 1,
              secinfo = list(header = make_header("14.1.2", "Age Summary (tfrmt)"),
                             footer = make_footer("gt_tbl", "tfrmt"))) |>
  rtf_tables(as_rtftables(tf_gt))

out <- tempfile(fileext = ".rtf")
generate_rtfreport(doc, out, overwrite = TRUE)
file.exists(out)
```

## 4. A subject listing (plain data.frame)

When you do not need a table-building package -- e.g. a simple listing --
hand rtfreporter a `data.frame`/tibble directly and set the display
metadata yourself:

```{r listing}
lst <- adsl |>
  arrange(USUBJID) |>
  transmute(USUBJID, Arm = ARM, Age = AGE, Sex = SEX) |>
  head(12)

doc <- rtf_document() |>
  rtf_section(page = 1,
              secinfo = list(header = make_header("16.2.1", "Subject Listing"),
                             footer = make_footer("data.frame / tibble", "base R"))) |>
  rtf_tables(as_rtftables(lst),
             col_header = c("Subject ID", "Arm", "Age", "Sex"),
             col_rel_width = c(3, 3, 1, 1))

out <- tempfile(fileext = ".rtf")
generate_rtfreport(doc, out, overwrite = TRUE)
file.exists(out)
```

## Assembling a multi-table deliverable

Real deliverables bundle many TLGs into one file.  Build each page's
content, then concatenate the RTFs with
[`assemble_rtf()`](../reference/assemble_rtf.html), or append several
tables to one document with successive `rtf_tables()` calls.

```{r assemble}
f1 <- tempfile(fileext = ".rtf"); f2 <- tempfile(fileext = ".rtf")
generate_rtfreport(
  rtf_document() |>
    rtf_section(page = 1, secinfo = list(header = make_header("14.1.1", "Demographics"),
                                         footer = make_footer("gt_tbl", "gtsummary"))) |>
    rtf_tables(as_rtftables(dm_tbl, align_count_pct = TRUE)),
  f1, overwrite = TRUE)
generate_rtfreport(
  rtf_document() |>
    rtf_section(page = 1, secinfo = list(header = make_header("14.3.1", "Adverse Events"),
                                         footer = make_footer("rtables TableTree", "tern + rtables"))) |>
    rtf_tables(as_rtftables(ae_tbl, align_count_pct = TRUE)),
  f2, overwrite = TRUE)

bundle <- tempfile(fileext = ".rtf")
assemble_rtf(c(f1, f2), bundle)
file.exists(bundle)
```

## Notes

* The output is `.rtf`; open it in Word / LibreOffice (or batch-convert to
  PDF).  See the homepage 30-second example for a screenshot of the
  rendered look.
* What does and does not carry across from each source is documented in
  the *What is carried, by source* table in
  [`?as_rtftables`](../reference/as_rtftables.html).
* The data here is simulated and the article is fully self-contained; no
  CDISC pilot data is bundled.
