---
title: "TLG cookbook: pharmaverse tables -> RTF"
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
.pkgs <- c("gtsummary", "rtables", "tern", "tfrmt", "random.cdisc.data",
           "dplyr", "tidyr")
.ok <- all(vapply(.pkgs, requireNamespace, logical(1L), quietly = TRUE))
knitr::opts_chunk$set(eval = .ok)
```

```{r, eval = TRUE, echo = FALSE, results = "asis"}
if (!.ok)
  cat("> This article renders only when gtsummary, rtables, tern, tfrmt,",
      "random.cdisc.data, dplyr and tidyr are installed.  The code below",
      "shows the workflow either way.\n")
```

rtfreporter does not compute statistics or format numbers -- the
[pharmaverse](https://pharmaverse.org) ecosystem already does that
beautifully.  rtfreporter handles the **last mile**: taking whatever table
object your team produced and laying it out as a clinical-style RTF page,
with the pagination, blank rows and running headers a real deliverable
needs.

This cookbook follows the
[pharmaverse TLG examples](https://pharmaverse.github.io/examples/tlg/):
the *Demographic* table is built **three ways** (tern, gtsummary, tfrmt)
and the *Adverse Events* table is paginated across several pages.  Each is
converted to RTF with [`as_rtftables()`](../reference/as_rtftables.html) +
[`rtf_tables()`](../reference/rtf_tables.html) +
[`generate_rtfreport()`](../reference/generate_rtfreport.html).

```{r setup, eval = TRUE}
library(rtfreporter)
```

```{r libs}
library(rtables)
library(tern)
library(gtsummary)
library(tfrmt)
library(dplyr)
library(tidyr)
```

## Data

We use the CDISC pilot ADaM data shipped with `random.cdisc.data` (the same
data the NEST TLG catalog uses), so the examples are realistic and fully
reproducible.

```{r data}
adsl <- random.cdisc.data::cadsl
adae <- random.cdisc.data::cadae
```

A reusable 3-row clinical header (company / protocol / centred table title)
and a one-line provenance footer:

```{r hf}
make_header <- function(table_no, title2) {
  rtf_header(rows = list(
    c(l = "Hoge Co. Limited",   r = "CONFIDENTIAL"),
    c(l = "Protocol: RTF-101",  r = "Page {AUTO_PAGE} of {AUTO_TOTAL_PAGES}"),
    c(c = paste0("Table ", table_no, "  ", title2))
  ))
}
make_footer <- function(source_object, built_with) {
  rtf_footer(c(l = paste0(
    "Source object: ", source_object, " (", built_with, "), ",
    "converted to RTF by rtfreporter.  ",
    "CDISC pilot data (random.cdisc.data); layout after the pharmaverse ",
    "TLG examples.")))
}
demog_title <- list(c("Demographic and Baseline Characteristics",
                      "Safety Analysis Set"))
```

## 1. Demographics, three ways

The same demographic summary, built with three different pharmaverse tools.
Whichever you prefer, `as_rtftables()` reads its object and the RTF comes
out the same clinical shape.

### 1a. tern + rtables

```{r dm-tern}
dm_lyt <- basic_table(show_colcounts = TRUE) |>
  split_cols_by("ARM") |>
  add_overall_col("All Patients") |>
  analyze_vars(
    vars   = c("AGE", "SEX", "RACE"),
    .stats = c("mean_sd", "median", "range", "count_fraction")) |>
  append_topleft("Characteristic")
dm_tern <- build_table(dm_lyt, adsl)
dm_tern
```

```{r dm-tern-rtf}
doc <- rtf_document() |>
  rtf_section(page = 1,
              secinfo = list(header = make_header("14.1.1a", "Demographics (tern)"),
                             footer = make_footer("rtables TableTree", "tern + rtables"))) |>
  rtf_tables(as_rtftables(dm_tern, blank_rows = "between_groups"),
             titles = demog_title)
out_tern <- tempfile(fileext = ".rtf")
generate_rtfreport(doc, out_tern, overwrite = TRUE)
file.exists(out_tern)
```

### 1b. gtsummary

```{r dm-gts}
dm_gts <- adsl |>
  select(ARM, AGE, SEX, RACE) |>
  tbl_summary(by = ARM,
              label = list(AGE = "Age (years)", SEX = "Sex", RACE = "Race"),
              statistic = list(all_continuous() ~ "{mean} ({sd})")) |>
  modify_header(label = "Characteristic") |>
  add_overall()
dm_gts
```

```{r dm-gts-rtf}
doc <- rtf_document() |>
  rtf_section(page = 1,
              secinfo = list(header = make_header("14.1.1b", "Demographics (gtsummary)"),
                             footer = make_footer("gt_tbl", "gtsummary::tbl_summary()"))) |>
  rtf_tables(as_rtftables(dm_gts, align_count_pct = TRUE,
                          blank_rows = "between_groups"),
             titles = demog_title)
out_gts <- tempfile(fileext = ".rtf")
generate_rtfreport(doc, out_gts, overwrite = TRUE)
file.exists(out_gts)
```

### 1c. tfrmt

`tfrmt` describes the table format declaratively (here: `xx.x` for the age
rows, a combined `n (xx.x%)` for the categorical rows) and renders to gt.

```{r dm-tfrmt}
age_long <- adsl |>
  group_by(ARM) |>
  summarise(Mean = mean(AGE), SD = sd(AGE), .groups = "drop") |>
  pivot_longer(c(Mean, SD), names_to = "label", values_to = "value") |>
  mutate(group = "Age (years)", param = tolower(label))

cat_long <- function(var, group_lbl) {
  adsl |>
    count(ARM, !!sym(var)) |>
    group_by(ARM) |>
    mutate(pct = n / sum(n) * 100) |>
    ungroup() |>
    pivot_longer(c(n, pct), names_to = "param", values_to = "value") |>
    mutate(group = group_lbl, label = as.character(.data[[var]])) |>
    select(ARM, group, label, param, value)
}
dm_long <- bind_rows(age_long |> select(ARM, group, label, param, value),
                     cat_long("SEX", "Sex"),
                     cat_long("RACE", "Race"))

tf <- tfrmt(
  group = group, label = label, column = ARM, param = param, value = value,
  body_plan = body_plan(
    frmt_structure("Age (years)", ".default", frmt("xx.x")),
    frmt_structure(".default",    ".default",
                   frmt_combine("{n} ({pct}%)",
                                n = frmt("xx"), pct = frmt("xx.x")))))
dm_tfrmt <- print_to_gt(tf, dm_long)
dm_tfrmt
```

```{r dm-tfrmt-rtf}
doc <- rtf_document() |>
  rtf_section(page = 1,
              secinfo = list(header = make_header("14.1.1c", "Demographics (tfrmt)"),
                             footer = make_footer("gt_tbl", "tfrmt"))) |>
  rtf_tables(as_rtftables(dm_tfrmt, blank_rows = "between_groups"),
             titles = demog_title)
out_tfrmt <- tempfile(fileext = ".rtf")
generate_rtfreport(doc, out_tfrmt, overwrite = TRUE)
file.exists(out_tfrmt)
```

## 2. Adverse events (paginated across pages)

A standard "AEs by SOC and Preferred Term" table.  With a real adverse-event
data set it is long, so we let `as_rtftables()` paginate it -- `split =
"group_safe"` keeps each SOC block whole, spilling to the next page rather
than breaking a group, and `blank_rows = "between_groups"` separates the
SOCs.  rtfreporter repeats the column headers and the running header on
every page automatically.

```{r ae-build}
ae_lyt <- basic_table(show_colcounts = TRUE) |>
  split_cols_by("ARM") |>
  analyze_num_patients(
    vars = "USUBJID", .stats = "unique",
    .labels = c(unique = "Total number of patients with at least one AE")) |>
  split_rows_by("AEBODSYS", label_pos = "topleft",
                split_label = "MedDRA System Organ Class") |>
  summarize_num_patients(var = "USUBJID", .stats = "unique",
                         .labels = c(unique = "Total patients with an AE")) |>
  count_occurrences(vars = "AEDECOD")
ae_tbl <- build_table(ae_lyt, adae, alt_counts_df = adsl)

ae_pages <- as_rtftables(ae_tbl,
                         split      = "group_safe",
                         max_rows   = 12,
                         blank_rows = "between_groups",
                         align_count_pct = TRUE)
length(ae_pages)            # number of RTF pages
```

```{r ae-rtf}
# One section starting at page 1 covers every page; rtfreporter repeats its
# header/footer (and the "Page X of Y" field) on each page automatically.
doc <- rtf_document() |>
  rtf_section(page = 1,
              secinfo = list(header = make_header("14.3.1",
                                                  "Adverse Events by SOC and Preferred Term"),
                             footer = make_footer("rtables TableTree", "tern + rtables"))) |>
  rtf_tables(ae_pages,
             titles = rep(list(c("Adverse Events Summary", "Safety Analysis Set")),
                          length(ae_pages)))
out_ae <- tempfile(fileext = ".rtf")
generate_rtfreport(doc, out_ae, overwrite = TRUE)
file.exists(out_ae)
```

## 3. Assembling a deliverable

A real submission bundles many TLGs into one file.  Render each table to its
own `.rtf`, then concatenate them with
[`assemble_rtf()`](../reference/assemble_rtf.html) (page numbering and
section headers continue across the seam):

```{r assemble}
bundle <- tempfile(fileext = ".rtf")
assemble_rtf(c(out_gts, out_ae), bundle)   # demographics + the multi-page AEs
file.exists(bundle)
```

## Notes

* The output is `.rtf`; open it in Word / LibreOffice (or batch-convert to
  PDF).
* `split = "group_force"` cuts strictly at `max_rows` (inserting `(Cont.)`
  rows), while `"group_safe"` never breaks a group; `"by_value"` puts one
  group per page.  See [`as_rtftables()`](../reference/as_rtftables.html).
* What does and does not carry across from each source is documented in the
  *What is carried, by source* table in `?as_rtftables`.
